---
title: "Data Visualization with ggplot"
author: "Zach Compton"
date: "10/3/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Example from tutorial 
Set your working directory to the file that contains your dataset
```{r}
setwd("~/Desktop/Comp Onc Projects/current big datasets")
```

Read in data sheet with it named as "Data"
```{r}
Data <- read.csv('ggplotData.csv')
```

What variables are included in the dataset?
```{r}
head(Data)
```

Now let's read in our tidyverse library. Tidyverse is a package that includes the ggplot data visualization toolkit. If you have never installed this package you can do so with the line
install.packages('tidyverse'). If you have already installed tidyverse you can simple load the library 
```{r}
library(tidyverse)
```

#### Basic Scatter Plot
Let's start with a bare bones plot of the relationship between adult_weight and max_longevity
```{r}
ggplot(Data, aes(x = max_longevity, y = adult_weight)) + 
  geom_point()
```

Yuck! Our data is super bunched up with a few outliers, making the graph difficult to interpret. Lets log10 transform both variables and see if we can rescue this plot.
```{r}
Data$max_longevity <- log10(Data$max_longevity)
Data$adult_weight <- log10(Data$adult_weight)
```

Retry the plot
```{r}
ggplot(Data, aes(x = max_longevity, y = adult_weight)) + 
  geom_point()
```

Much better! Now we can begin working through all the various arguments within ggplot to improve our visualization techniques so we can tell the most compelling story possible with our data.

Let's see this same graph but assigning a specific color for each of the three Class contained within our dataset
```{r}
ggplot(Data, aes(max_longevity, adult_weight, colour = Class)) + 
  geom_point()
```

What about adding a trend line. The argument added to the geom_smooth function instructs on how to build the line. method = "lm" creates a linear model, the line of best fit. Here we dropped the colour = Class argument because if we keep it there will be a line created for each Class as shown below.
```{r}
ggplot(Data, aes(max_longevity, adult_weight, colour = Class)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x)
```

The grey shaded area around each line is a confidence region. The confidence region can be interpreted as the defined space that the line is equally likely to occupy. Typically, confidence regions shrink with increases in data points. But if we want to plot multiple lines they can be distracting, let's remove them.
```{r}
ggplot(Data, aes(max_longevity, adult_weight, colour = Class)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, se= FALSE)
```


Let's return to a basic graph and walk through all of the ways to make aesthetics adjustments to our graphs

We can use size and color arguments to manually adjust the geom_smooth line. A list of available colors will be printed by using the colors() function. Size can be positive number but scales up rapidly. See the difference between 1.2 and 12.2
```{r, message=FALSE, results=FALSE, warning=FALSE}
ggplot(Data, aes(max_longevity, adult_weight)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE)
```


```{r}
ggplot(Data, aes(max_longevity, adult_weight)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 10.2, color = "salmon4", se = FALSE)
```

#### X and Y Axis Scaling 
The scale (range) of the X and Y axis is determined by the range of the data points but it can be manually adjusted through two methods. The first uses the xlim and ylim arguments.
```{r, warning=FALSE}
ggplot(Data, aes(max_longevity, adult_weight)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  xlim(c(0,4)) +
  ylim(c(0, 4))
```

Too zoomed out, but the inverse is almost worse. Vital to understand that ylim and xlim arguments will remove any of your data points that dont fall within the limit you have set.
```{r, warning=FALSE}
ggplot(Data, aes(max_longevity, adult_weight)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  xlim(c(1,3)) +
  ylim(c(1, 3))
```

In this plot the auto generated X and Y axis are fine. But lets add some X and Y axis labels that are prettier than the generic column names that are in place now.
```{r}
ggplot(Data, aes(max_longevity, adult_weight)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  ylab("Log10 Adult Body Mass (g)") +
  xlab("Log10 Maximum Longevity (m)")
```

If you haven't figured it out yet ggplot works by adding layers to plots. The layers are added together using a + sign. Let's add a layer to our graph with a main title.
```{r}
ggplot(Data, aes(max_longevity, adult_weight)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  ylab("Log10 Adult Body Weight (g)") +
  xlab("Log10 Maximum Longevity (m)") +
  ggtitle("Maximum Longevity vs. Adult Body Mass")
```

Adding axis titles, main titles, and subtitles can all be combined into one argument using the labs() function
```{r}
ggplot(Data, aes(max_longevity, adult_weight)) + 
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  labs(title="Maximum Longevity vs. Adult Body Mass", subtitle="Mammals, Birds, and Reptiles", y="Log10 Adult Body Weight (g)", x="Log10 Maximum Longevity (m)")
```

Now lets color our points by species. But first lets create a custom color palette for **spooky** season by c-binding some haunted hex codes together.

```{r}
spooky_palette <- rep(c("#F75F1C", "#FF9A00", "#000000", "#881EE4", "#85E21F"), length.out = 250)
```

```{r}
ggplot(Data, aes(max_longevity, adult_weight, colour = Species)) + 
  scale_color_manual(values=spooky_palette)+
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  labs(title="Maximum Longevity vs. Adult Body Mass", subtitle="Mammals, Birds, and Reptiles", y="Log10 Adult Body Weight (g)", x="Log10 Maximum Longevity (m)")
```

That's no good, with all the species we have the legend is too big. Let's drop the legend.
```{r}
ggplot(Data, aes(max_longevity, adult_weight, colour = Species)) + 
  scale_color_manual(values=spooky_palette)+
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  labs(title="Maximum Longevity vs. Adult Body Mass", subtitle="Mammals, Birds, and Reptiles", y="Log10 Adult Body Weight (g)", x="Log10 Maximum Longevity (m)") + 
   theme(legend.position = "none")
```

We can also remove the grey background
```{r}
ggplot(Data, aes(max_longevity, adult_weight, colour = Species)) + 
  scale_color_manual(values=spooky_palette)+
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  labs(title="Maximum Longevity vs. Adult Body Mass", subtitle="Mammals, Birds, and Reptiles", y="Log10 Adult Body Weight (g)", x="Log10 Maximum Longevity (m)") + 
  theme_bw(base_size = 15)+
   theme(legend.position = "none")
```

## Faceting

Faceting can be a brilliant way to visualize your data across "bins". Bins mean any categorical variable in which you can divide your data into. As we can see with this data a natural bin to use is Class. Now lets a make a different plot of longevity vs adult weight for each Class in a super simple way.
```{r}
ggplot(Data, aes(max_longevity, adult_weight, colour = Species)) + 
  scale_color_manual(values=spooky_palette)+
  geom_point() + 
  geom_smooth(method = "lm", formula = y ~ x, size = 1.2, color = "black", se = FALSE) +
  labs(title="Maximum Longevity vs. Adult Body Mass", subtitle="Mammals, Birds, and Reptiles", y="Log10 Adult Body Weight (g)", x="Log10 Maximum Longevity (m)") + 
   theme(legend.position = "none")+
  facet_wrap(~Class)
```

