---
title: "dplyr"
author: "Zach Compton"
date: "11/4/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load Libraries 
```{r}
install.packages('tidyverse')
library(tidyverse)
```
## Simulating data
Lets simulate data for 10 random animals that I am making up.
```{r cars}
Species <- c("chicken", "duck", "tiger", "unicorn", 
             "rabbit", "monky", "platypus", "rhino", "turkey", "hippo")
Lifespan <- rnorm(10, 56, 20)
BodyMass <- rnorm(10, 150, 40)
NeoplasiaPrevalence <- rnorm(10,15,5)
```

Combine these three created variables into a single dataframe
```{r}
Data <- data.frame(Species, Lifespan, BodyMass, NeoplasiaPrevalence)
```

We can check out our data with the head function
```{r}
head(Data)
```

You can also specify how many rows (observations) you want to see at one time. Here I will say 10 so that we can see our whole dataset
```{r}
print(Data, 10)
```

## Filter
You can use the filter function to search/subset columns by a specified criteria. Let's find species with a neoplasia prevalence less that 5%
```{r}
filter(Data, NeoplasiaPrevalence < 50)
```

We can add two conditional statements so that to be included in the subset the data need to meet to criteria (or as many criteria as you want to stack)
```{r}
filter(Data, NeoplasiaPrevalence > 5 , Lifespan < 50)
```

We can add an "or" statement to this by asking for NeoplasiaPrevalence that is also over 15%
```{r}
filter(Data, NeoplasiaPrevalence < 5 | NeoplasiaPrevalence > 15)
```

Quick note - we can also ask for "less than or equal to" or "greater than or equal to" as to not exclude species if they *happen* to be at exact value we ask for
```{r}
filter(Data, NeoplasiaPrevalence <= 5 | NeoplasiaPrevalence >= 15)
```

Of course if we are going to use this subset for further analyses we should variable name it so we can reference it in said analyses
```{r}
Data2 <- filter(Data, NeoplasiaPrevalence <= 5 | NeoplasiaPrevalence >= 15)
```

## Select 
The filter function is about selecting observations (rows) but what if we want to subset by entire variables (columns)?
Let's say we just wanted a data frame of only NeoplasiaPrevalence values
```{r}
Subset <- dplyr::select(Data, NeoplasiaPrevalence)
```

Notice the syntax here was a little different. I will show why in this example
```{r, echo=T}
dplyr::select(Data, NeoplasiaPrevalence)
```
This is a great example of a very common function not working even though it is set up correctly. 
There are seveal different 'select' functions in r so you can use the dyplr::select to specify which package, i.e. dplyr, the function you are trying to use is from.

## Mutate
What if we want to create a new variable (column) in your data set, like we often have to do when we want to combine several life history variables
```{r}
Data <- mutate(Data, BodyMassxLifespan = BodyMass * Lifespan)
```

## Piping functions together
When you want to be able to execute multiple functions at one you can tie them together with a pipe function %>%
```{r}
Data %>%
   dplyr::select(BodyMass) %>%
  filter(BodyMass > 50) 
```

## Arrange
We can also arrange our data by certain criteria, like ascending or descening in value, much like an excel filter
```{r}
Data %>%
  dplyr::select(BodyMass, Lifespan) %>%
  arrange(desc(BodyMass))
```

## Summarise
The summarise function (with a S, not a Z) is very similar to the base R 'tapply' function that allows us to execute some statistical summary on multiple columns at once
```{r}
Data %>% 
  summarise_each (funs(mean), BodyMass, Lifespan, NeoplasiaPrevalence)
```

For those of you managing large comparative pathology data sets one useful function you could tack on the end of a series of searches is the n_distinct function which counts the number of distinct entries in a given column
```{r}
Data %>%
n_distinct(Species)
```

## Random useful things
You can take a random sample of observations (rows) from your data set. The first way is by take a fixed number sample 
```{r}
Data %>% sample_n(1000)
```
Or by specifying a proportion of rows given your total data set
```{r}
Data %>% sample_frac(0.25)
```
```{r}
install.packages("tmaptools")
library(tmaptools)
palette_explorer() 
```

