---
title: "Sex Bias Violins"
author: "Zach Compton"
date: "12/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
#### Libraries
```{r, message = F}
library(tidyverse)
library(cowplot)
```

### Make Violins 

Read data in. This data set has been specifically designed to be used just for this violin script.
Look at the head of the data to get a sense of how a data frame needs to be arranged for violin plots
```{r}
setwd("~/Desktop/Git/sexBias")
Data <- read.csv("malignancy_combined.csv")
head(Data)
```

Subset by class
```{r}
Mammals <- Data[Data$Clade=="Mammalia", 
                    c("Prevalence", "Sex", "Clade")]
Amphibians <- Data[Data$Clade=="Amphibia", 
                    c("Prevalence", "Sex", "Clade")]
Reptiles <- Data[Data$Clade=="Reptilia", 
                    c("Prevalence", "Sex", "Clade")]
Aves <- Data[Data$Clade=="Aves", 
                    c("Prevalence", "Sex", "Clade")]

```

Violin code
Note-- The stats included in this script come from a Fisher's exact test 
```{r}
mam_vio <- ggplot(Amphibians, aes(x=Sex, y=100*Prevalence, fill=Sex)) + 
  geom_violin(adjust=1) +
  theme(
    legend.title = element_text(size = 21, face = "bold"),
    legend.text = element_text(size = 20)
  ) +
 labs(title ="Malignancy Prevalence by Sex in Amphibians", subtitle = "Fisher's Exact Test: P-Value = 0.7796, Odds Ratio =0.72") +
  labs(fill = 'Sex')+
ylab("Malignancy Prevalence %") +
  xlab("Sex") +
  theme(
    plot.title = element_text(size = 20, face = "bold")) +
  theme(axis.text=element_text(size=16),
        axis.title=element_text(size=16)) 
print(mam_vio)

```

Add some mean bars and clean up the background with cowplot 
```{r}
mam_vio + geom_jitter(shape=16, position=position_jitter(0.2))+
stat_summary(fun=mean, geom="crossbar", size=0.7) + 
scale_fill_manual(values = c("#56B4E9", "#CC79A7", "#F0E442")) +
theme_cowplot(12) 
  theme(legend.position = "none")

```

### Density Plots 

Single Density Plot
```{r}
ggplot(Data, aes(x=Prevalence*100,y=..density..,fill=factor(Sex))) + 
  geom_density(alpha=0.25) +
  labs(title  = "Distribution of Malignancy Prevalence by Sex in Aves",fill = "Sex")+
  scale_fill_discrete(labels = c("Male", "Female"))+
  theme_cowplot(12)
```

Install packages for aligning ggplot graphs together
```{r, message=F, error=F}
library("gridExtra")  
```

Make and NAME the plots!!

#### Neoplasia Distribution
```{r}

mam<-ggplot(Mammals, aes(x=Prevalence*100,y=..density..,fill=factor(Sex))) + 
  geom_density(alpha=0.25) +
  labs(title  = "Mammalia",fill = "Sex")+
  xlab("Malignancy Prevalence") +
   ylab(NULL)+
  scale_fill_discrete(labels = c("Male", "Female"))+
  theme_cowplot(12)

rep<-ggplot(Reptiles, aes(x=Prevalence*100,y=..density..,fill=factor(Sex))) + 
  geom_density(alpha=0.25) +
  labs(title  = "Reptilia",fill = "Sex")+
  xlab("Malignancy Prevalence") +
   ylab(NULL)+
  scale_fill_discrete(labels = c("Male", "Female"))+
  theme_cowplot(12)

amph<-ggplot(Amphibians, aes(x=Prevalence*100,y=..density..,fill=factor(Sex))) + 
  geom_density(alpha=0.25) +
  labs(title  = "Amphibia",fill = "Sex")+
  xlab("Malignancy Prevalence") +
   ylab(NULL)+
  scale_fill_discrete(labels = c("Male", "Female"))+
  theme_cowplot(12)

aves<-ggplot(Aves, aes(x=Prevalence*100,y=..density..,fill=factor(Sex))) + 
  geom_density(alpha=0.25) +
  labs(title  = "Aves",fill = "Sex")+
  xlab("Malignancy Prevalence") +
  ylab(NULL)+
  scale_fill_discrete(labels = c("Male", "Female"))+
  theme_cowplot(12)
```

String them together into one plot
```{r}
grid.arrange(mam, rep, aves, amph, ncol = 2)         
```

